Under-ice acoustic navigation using real-time model-aided range estimation

### 2022-02-21 rebuttal/rebuttal-letter/trunk-rebuttal.tex

\documentclass{article}
	\def\papertitle{Under-ice acoustic navigation using real-time model-aided range estimation}
	\def\authors{E. Bhatt, O. V\'{i}quez, H. Schmidt}
	\def\journal{Journal of the Acoustical Society of America}
	\def\doi{}

@@

For clarity, we also more prominently establish definitions for navigation, positioning, and re-positioning to better contextualize our results and contributions, based on this discussion in Howe et al (2019) \emph{Observing the Oceans Acoustically} (and many others):

\begin{quote} 
PNT is a combination of three distinct, constituent capabilities:

\begin{enumerate}
	\item Positioning is the ability to accurately and precisely determine one’s location referenced to a standard geodetic system;
	\item Navigation is the ability to determine current and desired position (relative or absolute) and apply corrections to course, orientation, and speed to attain a desired position anywhere in the domain of concern; 
	\item Timing is the ability to acquire and maintain accurate and precise time anywhere in the domain of interest within user-defined timeliness parameters; it also includes time transfer.
\end{enumerate}

Navigation is real-time and necessarily depends on positioning, and positioning depends on timing.
\end{quote}

Trilateration of beacon events was done in real-time, on the beacons themselves.
In this work, we further delineate that positioning is in real-time and re-positioning is post-processed.
We have also added re-positioning and re-navigation trilateration results to emulate our system performance given the navigation results have no ground truth GNSS to compare to.

@@

%% ------- COMMENTS FROM REVIEWER ONE ------ %%
\section{Reviewer One}

	\subsection*{Summary}

	\ebline{This paper demonstrates how a positioning system over \texttt{\small $\sim$}1.5-3 km of range achieves 11 m of positioning uncertainty. The work demonstrates how a single group velocity calculated from acoustic rays predicted by BELLHOP can be used to estimate ranges to be used in a LBL network for positioning. The data presented are from transceivers on fixed ice-tethered buoys but the system could also be implemented on an autonomous underwater vehicle.}

	This summary captures key results presented in the paper. However, we note that the system was successfully deployed on an underwater vehicle during ICEX20. The results from this work are specifically limited to the methodology behind calculating a single group velocity, which we attempt to validate through beacon-to-beacon communications with known GNSS timestamps.

@@

	\subsection*{Overall Comments}

	\reviewer{}{The claims made in the abstract need to be better supported by the data presented or else tempered a bit. What is demonstrated here is not real-time navigation on a vehicle, but calculation of a handful of short fixed ranges from ice-tethered buoys. The limited data presented here begs more discussion with regards to how applicable the algorithm would be in normal vehicle operation and to navigation of the vehicle, as there were limited ranges between the beacons. What depths does the vehicle typically operate at and what is the maximum range from the beacons?  The abstract stresses real-time results, but the real-time results are just 3 paragraphs, and the paper mostly concentrates on post-processing results.} \label{1dot1}

	We have adjusted the claims made in the abstract (and throughout the paper) to better emphasize what was deployed in the field, as part of normal vehicle operations, and how the proposed improvements compare to the fielded approach. We have presented data that we feel better support the claims about single meter accuracy as well. 
	While we initially aimed to contextualize the calculation of a handful of short fixed ranges between ice-tethered buoys as the proxy for navigation, we agree with the reviewer that this representation of our work is incomplete in portraying its portability to vehicle operations. Accordingly, we added the same analysis for the AUV run in ICEX20.

@@

	Thank you for the suggestion to discuss in terms of range, vehicle, and propagation conditions. We have not done this explicitly but taken care to illustrate the differences, and what parts we are comparing.

	Lastly, yes, we did mean real-time navigation, but have removed the line in question.

	\reviewer{}{Related to the point above, the introduction should make clear what type of vehicles the proposed method of navigation is applicable to and over what ranges. The authors do start to do this when they note that a shallow receiver depth enables the estimation method (line 274), but should more directly specify under what conditions the proposed algorithm is applicable.  In the Discussion section, there is an implication that this method of using a single group velocity could be applied to mesoscale operations but should acknowledge that propagation at these larger scales results in arrivals that are spread over several seconds including multiple modal arrivals with varying group velocities.  This method would presumably result in errors much higher than 11m….how might that scale?}

	Within the data collected during ICEX20, we actually do see instances of multiple arrivals detected at a single receiver, spread over multiple seconds\textemdash this can be exemplified by arrivals consistent with bottom-bounce paths.

	The point of the Nearest Bounce Criteria, the improved approach proposed in this paper as part of the post-processing analysis, is precisely to expose how tracking the results of the acoustic model by modal variations can impact the ranging and positioning results; this is done using field data as the foundation. In our case, at short ranges under the Arctic ice, we have opted to track boundary interactions as the basis for modal variation; however, a different arrival classification basis may be needed for longer ranges in the SOFAR channel, for example.

@@

	\reviewer{}{I understand what the authors are referencing when they use the term “group velocity” but I think this term should be used a bit more carefully.  Group velocities are often associated with modal propagation in which each mode has a distinct group velocity.  The goal in this work is to determine a single group velocity by taking a power weighted average of several different ray arrivals to derive range.  This is addressed somewhat in lines 395 and following, but I think it would be appropriate to expand on this discussion and move it up.  Lines 235 and following state “the variation of horizontal group velocity from any source-receiver pair is the fundamental challenge to implement GPS-like navigation in an LBL navigation paradigm…” this statement could use some clarification.  Also, perhaps change the axis on Figure 5 to read something like ‘implied group velocity.”} \label{1dot5}

	Thank you, these are all excellent suggestions. Instead of ``group velocity'' we use ``effective sound speed'', noted by $c$. The statement in question has been rewritten to talk about a single sound speed compensating for ray refraction and reflection, and is on line \ref{v1:1.5}.

@@

	Thank you. We have included this point and make the distinction that RAFOS floats are re-positioned.

	\lreviewer{line 192}{v1:1.12}{Clarify what you mean when you state that the transmit and receive layers are independent? Also, the discussions of physical and virtual layers are confusing. I think everything is there, but reorganization would make this more clear. Consider moving up the paragraph starting at line 203.}

	We have removed the jargon of the physical and virtual layers in favor of a more streamlined explanation. The transmit and receive layers are activated independently.

	\lreviewer{line 205}{v1:1.13}{What do you mean that the coordinate system is range independent?}
	We meant that the analysis of beacon-to-beacon connections is presented in r/z, not x/y/z. We found this framing awkward as it is more common to say the propagation has a depth dependent sound speed profile. Range-dependent models may assume no change in the sound speed profile, but account for a range dependent bathymetry; or they may account for changes in both bathymetry and SSP over range. This work includes a range dependent bathymetry.

@@

	\lreviewer{line 222}{v1:1.14}{Could you provide more information on the baseline and weighted SSPs. Where did the baseline profile come from?}
	\label{1dot14}

	In anticipation of ICEX20, all of the ITP data for the Beaufort Sea was assimilated. For operation reasons, we designed a system to transmit sound speed updates through digital acoustic message. This system relied on a simple basis function representation of the sound speed and its perturbations. Given the wide range of observed ducted conditions across all ITP data, several basis sets were created, corresponding to March of different years. During ICEX20, the data from Mar 2013 was found to be the closest match for observed conditions. The baseline is the mean sound speed from Mar 2013. The weighted SSP is the basis set projection for the CTD taken during ICEX20.

@@

	\lreviewer{Figure 1}{v1:fig:vehicleRecovery}{Figure 1 is a nice picture showing the vehicle, which motivates the work presented here, but was not used for collection of any of the data presented.  As it’s not pertinent to the data presented here, I’d suggest removing this figure or at least combining this figure with Figure 2.}

	Thank you. We have kept this image but added the vehicle track, as it is necessary for the re-navigation analysis as mentioned in Comment \ref{1dot1}.

	\lreviewer{Figure 3}{v1:fig:overview}{Figure 3 tallies the number of successful events, and I think the use of the directional arrows communicates the data well.  How many events are successful in comparison with those that are considered unsuccessful and what generally causes an event to be deemed unsuccessful?}

	Thank you. We have added a successful decoding percentage, defined as the ratio of decoded over detected messages. We have also added the total number of transmissions, so one can see the ratio of detected over transmitted. An event is deemed unsuccessful given any error in the transmitted headers or any data frame, even if the error has nothing to do with the information necessary for positioning.

@@

	Thank you. We hope this manuscript would be of relevance to the Arctic Acoustics special issue given two topics of interest: under-ice navigation systems for floats, gliders, and autonomous underwater vehicles, and comparisons of current Arctic acoustic conditions with historical data.

	\reviewer{}{Is the manuscript a clear, concise, reasonably self-contained presentation of the material, giving adequate references to related work? Is English satisfactory? Please indicate needed changes in your report.}
	{\textbf{Reviewer 1:} The English is satisfactory.  This paper could benefit from some reorganization and careful editing. Several references to figures and tables were missing or incomplete. The hypothesis was presented in the discussion, but could be improved if this was more clearly stated at the outset.  Please see attached report for specific comments and suggestions.

	\textbf{Reviewer 2:} Overall the language is fine, details below.}

	Thank you. We have significantly reorganized the paper to improve clarity as suggested.

@@

	\textbf{Reviewer 2:} NA}

	\reviewer{}{Does the paper make effective use of journal space, or are parts unnecessary, unimportant, or subject to condensation? If so, which?
	}
	{\textbf{Reviewer 1}: I'd suggest condensing the section on GPS drift (Section III E).

	\textbf{Reviewer 2}: The GPS section seems out of place. Can it be shortened to a few sentences and references?}

	We have significantly condensed the background for this section. In addition, we narrowed the discussion to the how trilateration error compares to that of GNSS at high latitudes and our particular GNSS unit.

### 2022-02-21 manuscript/main.tex
oscar feedback pt 3(?) and dropping c from NBC/MBC

\section{Introduction}
\label{sec:1}  
Autonomous underwater vehicles (AUVs) are increasingly capable platforms to explore and sample the ocean, particularly for remote and/or dangerous regions.
However, navigation uncertainty is a major challenge in considering AUVs as standard tools for oceanographic research.
While land and air-based robots utilize information from Global Navigation Satellite Systems (GNSS) to achieve stunning location accuracy and precision throughout the duration of their missions, AUVs cannot access GNSS fixes while underwater.
Therefore, underwater vehicles have relied on any combination of dead reckoning, hydrodynamic models, inertial navigation systems, doppler velocity logs, and acoustic baseline positioning systems for navigation \citep{paull_auv_2014}.
Limiting navigation error and drift requires an AUV to periodically stall on the surface and obtain a GNSS fix to reset its position error.
This foolproof method of self-positioning is undesirable for stealth, adverse weather conditions, and mission efficiency, and inaccessible in a GNSS-denied situation like an ice-covered environment.

% introduce LBL and precise statement of work
Of underwater acoustic navigation systems, long baseline (LBL) is the most GPS-like in style and scale, and most appropriate for mitigating drift without overburdening computation or payload size on the vehicle \citep{van_uffelen_global_2021}.
The state-of-the-art for LBL outsources depth to a pressure sensor and solves the two-dimensional localization problem with an isovelocity, linear scaling between one way travel time (OWTT) and range \citep{Eustice2006,Eustice2007,webster_advances_2012,Webster2009}.
This assumption is valid for short scale operations but oversimplifies propagation for larger and/or complex acoustic environments.
To achieve single meter, GNSS-like performance in a GNSS-denied environment, we demonstrate an embedded ray-based data processing algorithm to convert recorded OWTTs into pseudorange estimates.
This methodology was integrated onto the AUV \emph{Macrura}, deployed and recovered in the Beaufort Sea, in March 2020, during the Ice Exercise 2020 (ICEX20).
A physics-driven methodology that received an \textit{in situ} sound speed profile (SSP) was necessary despite the small operational domain because of the relatively high-risk mission environment\textemdash total under-ice conditions and a variable double ducted acoustic environment.

@@

Localization algorithms that do consider environmental or acoustic uncertainty tend to focus on longer and larger experiments, where spatio-temporal variability cannot be ignored.
These methods have also been reserved for post-processing as they can be labor intensive, computationally heavy, and/or require additional information like contemporaneous data.
For example, gliders navigating with kinematic flight models and equipped with acoustic modems were later unambiguously associated with predicted ray arrivals, resulting in roughly a kilometer error and hundred meters uncertainty over basin scale propagation \cite{VanUffelen2013}.
A follow up study investigated how a single temporally and spatially averaged SSP could mitigate position error for a four month glider mission \cite{vanuffelen2016localization}.
\citet{Wu2019} cross correlate three days of real acoustic records with synthetic ones generated through ocean model snapshots from HYCOM \cite{chassignet_hycom_2007}.
While potentially applicable for various ocean states, this is reliant on model realism and impractical for real-time operations.
A ``cold start'' algorithm that does not require prior knowledge of track, position, or sound speed information inputs a four-dimensional ocean model, constrained by tomography data, into a range dependent ray code to isolate the last path detected in a full multipath pattern \citep{Mikhalevsky2020}.
Then, a representative group speed is solved for alongside position in a least squares fashion. 
This approach is able to re-position a floating hydrophone array with an error of 58 m and a standard deviation of 32 m based on six sources 129--450 km away but remains to seen for real-time navigation.

@@

\llabel{2.1a}
Given that a lens introduces significant ray refraction, the Beaufort Lens is a shorthand for the spatio-temporal variability of the local temperature and sound speed maxima generally around 50 to 60 m in depth.
\llabel{2.4}
A neutrally buoyant layer of warm Pacific Summer Water creates a unique double ducted environment \textemdash the upper duct degrades signal coherence due to intensified ice interaction and the lower duct effectively traps sound for long range propagation \citep{Poulsen2016}. 
Modeling output \citep{duda_long-range_2019,duda_effects_2021} and experimental observations \citep{Ballard2020,Bhatt2021} suggest that, in the Beaufort Sea, the duct is persistent and widespread but not necessarily continuous; it and its acoustic effects can be non-existent, minimal, or drastic.
Transmissions in the upper duct, between the surface ice and the lens, experience minimal attenuation but degrade in signal coherence with repeated reflections under the ice.
In lower duct, between the lens and its conjugate depth in the Atlantic water, roughly 200 m, sound above 350 Hz is trapped near losslessly for long range propagation \citep{Poulsen2016}.

The Arctic, while remote, is the perfect place to demonstrate mature navigation technologies in real GNSS-denied conditions.
Thorough reviews of uncrewed vehicle operations in polar environments can be found in \citet{norgren2014unmanned} and \citet{Barker2020}; there is no comparable work in the Arctic for a short range AUV deployment in the Beaufort Lens.
Seminal \citep{brooke1981arcs,jackson1983autonomous,light1989autonomous,bellingham1995auv,hayes2002determining} and more recent AUV deployments \citep{,jakuba_long-baseline_2008,Kunz2008,Kukulya2010,Plueddemann2012,timmermans2013scales,fossum2021adaptive} witnessed the classical upward refracting sound speed profile that is amenable to an isovelocity assumption.

Of note, despite different platforms and scales, are recent glider deployments in the Canada Basin.
In 2014, in partially ice-covered conditions, a long range LBL system with WHOI Micro-Modems at 100 m exploited the lower duct for long range communication with two gliders \citep{Freitag2016,Webster2015}.
The sound speed value measured at the time of reception was used to estimate pseudorange in post-processing.
The beacon-to-beacon performance was excellent, achieving contact at ranges greater than 200 km with a position uncertainty of 40 m.
The beacon-to-glider performance, however, deteriorated due to missed contacts outside the duct, and was not described quantitatively.
In 2017, gliders were deployed in a region with no ice coverage \citep{Graupe2019}.
Ranges were linearly scaled by a statistical description of sound speed observations taken during the experiment, 1450 $\pm$ 6.5 m/s.
This resulted in an error of 550 m, which was reduced by a factor between 4 and 5, depending on the dive, using a post-processing acoustic arrival matching method.
Both cases exploit the lower duct for high fidelity communication at long ranges.
Unintuitively, the smaller nature of our deployment during ICEX20 is not a simplifying factor.
For source depths typical to vehicle operations, 30 to 200 m, a shadow zone spans from 2 to 6 kilometers in range \citep{schmidt_acoustic_2016}.

Compared to the previous small scale navigation efforts, the approach in this paper integrates real-time model-aided data processing to estimate a representative sound speed along a path from source to receiver, leveraging climatology, \textit{in situ} data, and fast acoustic modeling.
The paper is organized as follows.
Section \ref{sec:icex20} details the experimental approach and conditions during ICEX20.
Given that there is no GNSS ground truth for the vehicle position while underway, we first evaluate the real-time ranging performance of GPS-linked beacon-to-beacon communication events in section \ref{sec:realtime}.
Section \ref{sec:post} uses insights from field data to introduce a new ray filtering algorithm to improve range estimation.
Section \ref{sec:trilat} emulates the real-time processing pipeline to re-position beacon-to-beacon events and re-navigate AUV \emph{Macrura}.

@@

The results from this paper derive from data taken while deploying the AUV \emph{Macrura}, a custom Bluefin-21, during the Ice Exercise 2020 (ICEX20), in the Beaufort Sea, from March 8th to 11th.
The AUV deployment was supported by the Integrated Communication and Navigation Network (ICNN) \citep{schneider_self-adapting_2020,Randeni2020,randeni_high-resolution_2021} a specialized implementation of the LBL solution.
The ICNN was initially developed via numerous virtual experiments to ensure robust algorithms and interfaces between different hardware components. 
The simulation capabilities are largely physics-driven with a modular system of systems approach\textemdash an environmental simulator with sub-components for the ocean, including Arctic ice drift and ocean acoustic propagation; a vehicle simulator with sub-components for vehicle dynamics and navigation; a topside hardware simulator and acoustic communications simulator, both with a software-only configuration and a hardware-in-the-loop version \citep{schneider_netsim_2018}.
The virtual environment similarly emulates the interfaces between the real components to test the entire software pipeline.

@@

The ICNN is comprised of four ice buoys in a loose rectangle, roughly 2 km away from a central ice camp with a topside computer, as shown in Fig.~\ref{fig:icnnOverview}.
Each ice buoy is outfitted with a Garmin GPS 18x, with a pulse-per-second rising edge aligned to 1 microsecond and a spec sheet accuracy of 3 m, 95\% of the time.
The AUV and each ice buoy are outfitted with a WHOI Micro-Modem \citep{Singh2006}, with a four-element receiver array, a single transmitter, and one-tenth of a millisecond resolution.
\llabel{1.6a} Acoustic messages were sent with a 10 kHz carrier frequency, 5 kHz bandwidth, and phase-shift keying (PSK) modulation on a time-division multiple access schedule with a thirty-second cycle, giving room for two-way communication throughout the mission volume.
The receive and transmit elements were split between shallow and deeper depths\textemdash 30 and 90 m\textemdash to provide better coverage across the shadow zone.
While each buoy only has one transmit depth, all buoys have both receive depths but the active receive layer is consistent across all buoys. \llabel{1.12}
The design of the ICNN enables a self-adapting network to transmit and receive at the optimal depth to maintain contact with the AUV \citep{schneider_self-adapting_2020}.
The buoys do not encompass the full horizontal range of the vehicle but are positioned to minimize overlap in trilateration for spherical positioning \citep{deffenbaugh1996posit}.

@@

In the face of GNSS-denied navigation, this approach was anecdotally successful, as shown in Fig. \ref{fig:vehicleRecovery}.
The AUV \emph{Macrura} was deployed through a hydrohole from an ice camp but recovered through an emergency hydrohole.
A random disk error stalled \emph{Macrura} underneath the ice but did not prevent it from transmitting its location.
Due to an incoming storm, a team placed a physical marker on the ice at the location.
Three days later, \emph{Macrura} was found within a meter of the marker.
We view the emergency recovery as qualitative proof of the robustness of this navigation approach.
Nonetheless, this paper specifically addresses the third and fourth steps\textemdash the conversion of travel times into pseudoranges and its effect on trilateration.
By focusing on pseudorange estimates between GPS-tracked beacons, and re-running the trilateration pipeline, the results are decoupled from all other mechanisms in the ICNN.

@@

An important component to our navigation solution is an accurate estimation of a representative SSP for the ocean volume.
Previous field experience, during the Ice Exercise 2016 (ICEX16), demonstrated the negative effects of the Beaufort Lens on tracking and communication \citep{schmidt_acoustic_2016}.
Fig. \ref{fig:sspExpectation} shows historical, modeled, and \textit{in situ} sound speed data for both ICEX16 and ICEX20.
These three input streams were selected to mirror the information available on a submarine (personal conversation with LT B. Howard and LT CDR D. Goodwin). \llabel{1.8}
In the field, the SSP information was shared with the vehicle via basis representation compression on a lightweight digital acoustic message \citep{bhatt_embedded_2022}. \llabel{2.7}
All modeled data comes from HYCOM \cite{chassignet_hycom_2007}, which does not seem to capture the forcing mechanisms that cause the Beaufort Lens.
For ICEX16, the data-driven profile was sourced from nearby Ice Tethered Profilers (ITP) after the field experiment \cite{Krishfield2008,toole_ice-tethered_2011} and exhibits a fairly low lens; the historical profile is from the World Ocean Atlas.
For ICEX20, the chosen weights (data-driven) profile derives from an estimate of initial CTD casts taken on site, showing an intense warm water intrusion; the baseline (historical) profile, showing moderate ducted conditions, comes from the average of March 2013 ITP data.
This month best matched sea ice and sound speed conditions at the beginning of ICEX20 \citep{bhatt_embedded_2022}.
It is important to note that all profiles that do show the Beaufort Lens do so with different local sound speed maxima at different depths, reflective of the wide range of lens properties observed for all ITP data in the region. \llabel{2.5}
The variability of the lens height and prominence is the main reason an updatable SSP was integrated into the ICNN solution.

@@

During ICEX20, the HYCOM profile was available but never deployed. \llabel{1.14}
For post-processing comparison, we introduce both the HYCOM profile and an isovelocity case, 1441.8 $\pm$ 3.7 m/s, as the mean and standard deviation of the observed sound speed profile over the first 200 m.
This is a contrived value taken in the style of \citet{Graupe2019} for the sake of comparison; the default value in the LAMSS simulator, which was not environmentally informed nor used during ICEX20, was 1430 m/s.

@@

\subsection{Minimal bounce criteria (MBC)}

\llabel{1.5} The fundamental challenge to implement GNSS-like navigation, especially in an acoustically complex propagation environment, is characterizing a single sound speed to compensate for the effects of ray refraction and reflection.
The use of the acoustic modem network for tracking relies on the accurate estimates of travel times between the submerged platform and LBL beacons, supported by clock synchronization and a pre-determined scheduling of acoustic events.
\llabel{2.9} For the Beaufort Lens in particular, the strong multipath effects make it virtually impossible to deterministically predict the modem's detected arrival time.

>>>>

\section{Introduction}
\label{sec:1}  
Autonomous underwater vehicles (AUVs) are increasingly capable platforms to explore and sample the ocean, particularly for remote and/or dangerous regions.
However, navigation uncertainty is a major challenge in considering AUVs as standard tools for oceanographic research.
While land and air-based robots utilize information from Global Navigation Satellite Systems (GNSS) to achieve single-meter location accuracy and precision throughout the duration of their missions, AUVs cannot access GNSS fixes while underwater.
Therefore, underwater vehicles have relied on any combination of dead reckoning, hydrodynamic models, inertial navigation systems, doppler velocity logs, and acoustic baseline positioning systems for navigation \citep{paull_auv_2014}.
Limiting navigation error and drift requires an AUV to periodically stall on the surface and obtain a GNSS fix to reset its position error.
This foolproof method of self-positioning is undesirable for stealth, adverse weather conditions, and mission efficiency, and inaccessible in a GNSS-denied situation like an ice-covered environment.

% introduce LBL and precise statement of work
Of underwater acoustic navigation systems, long baseline (LBL) is the most GPS-like in style and scale, and most appropriate for mitigating drift without overburdening computation or payload size on the vehicle \citep{van_uffelen_global_2021}.
The state-of-the-art for LBL outsources depth to a pressure sensor and solves the two-dimensional localization problem with an isovelocity, linear scaling between one way travel time (OWTT) and range \citep{Eustice2006,Eustice2007,webster_advances_2012,Webster2009}.
This assumption is valid for short scale operations but oversimplifies propagation at larger scales or complex sound speed structure.
To achieve single meter, GNSS-like performance in a GNSS-denied environment, we demonstrate an embedded ray-based data processing algorithm to convert recorded OWTTs into pseudorange estimates.
This methodology was integrated onto the AUV \emph{Macrura}, deployed and recovered in the Beaufort Sea, in March 2020, during the Ice Exercise 2020 (ICEX20).
A physics-driven methodology that received an \textit{in situ} sound speed profile (SSP) was necessary despite the small operational domain because of the relatively high-risk mission environment\textemdash total under-ice conditions and a variable double ducted acoustic environment.

@@

Localization algorithms that do consider environmental or acoustic uncertainty tend to focus on longer duration and larger range experiments, where spatio-temporal variability cannot be ignored.
These methods have also been reserved for post-processing as they can be labor intensive, computationally heavy, and/or require additional information like contemporaneous data.
For example, gliders navigating with kinematic flight models and equipped with acoustic modems were later unambiguously associated with predicted ray arrivals, resulting in roughly a kilometer error and hundred meters uncertainty over basin scale propagation \cite{VanUffelen2013}.
A follow up study investigated how a single temporally and spatially averaged SSP could mitigate position error for a four month glider mission \cite{vanuffelen2016localization}.
\citet{Wu2019} cross correlate three days of real acoustic records with synthetic ones generated through ocean model snapshots from HYCOM \cite{chassignet_hycom_2007}.
While potentially applicable for various ocean states, this is reliant on model realism and impractical for real-time operations.
A ``cold start'' algorithm that does not require prior knowledge of track, position, or sound speed information inputs a four-dimensional ocean model, constrained by tomography data, into a range dependent ray code to isolate the last path detected in a full multipath pattern \citep{Mikhalevsky2020}.
Then, a representative group speed is solved for alongside position in a least squares fashion. 
This approach is able to re-position a floating hydrophone array with an error of 58 m and a standard deviation of 32 m based on six sources 129--450 km away but remains to seen for real-time navigation.

@@

\llabel{2.1a}
Given that a lens introduces significant ray refraction, the Beaufort Lens is a shorthand for the spatio-temporal variability of the local temperature and sound speed maxima generally around 50 to 60 m in depth.
\llabel{2.4}
A neutrally buoyant layer of warm Pacific Summer Water creates a unique double ducted environment \textemdash the upper duct degrades signal coherence due to intensified ice interaction and the lower duct effectively traps sound for long range propagation \citep{Poulsen2016}. 
Modeling output \citep{duda_long-range_2019,duda_effects_2021} and experimental observations \citep{Ballard2020,Bhatt2021} suggest that, in the Beaufort Sea, the duct is persistent and widespread but not necessarily continuous; it and its acoustic effects can be non-existent, minimal, or drastic.
Transmissions in the upper duct, between the surface ice and the lens, experience minimal attenuation but degrade in signal coherence with repeated reflections under the ice.
In the lower duct, between the lens and its conjugate depth in the Atlantic water (roughly 200 m), sound above 350 Hz is trapped near losslessly for long range propagation \citep{Poulsen2016}.

The Arctic, while remote, is the perfect place to demonstrate mature navigation technologies in real GNSS-denied conditions.
Thorough reviews of uncrewed vehicle operations in polar environments can be found in \citet{norgren2014unmanned} and \citet{Barker2020}; there is no comparable work in the Arctic for a short range AUV deployment in the Beaufort Lens.
Seminal Arctic AUV deployments \citep{brooke1981arcs,jackson1983autonomous,light1989autonomous,bellingham1995auv,hayes2002determining} and more recent ones \citep{,jakuba_long-baseline_2008,Kunz2008,Kukulya2010,Plueddemann2012,timmermans2013scales,fossum2021adaptive} witnessed the classical upward refracting sound speed profile that is amenable to an isovelocity assumption.

Of note, despite different platforms and scales, are recent glider deployments in the Canada Basin.
In 2014, in partially ice-covered conditions, a long range LBL system with WHOI Micro-Modems at 100 m depth exploited the lower duct for long range communication with two gliders \citep{Freitag2016,Webster2015}.
The sound speed value measured at the time of reception was used to estimate pseudorange in post-processing.
The beacon-to-beacon performance was excellent, achieving contact at ranges greater than 200 km with a position uncertainty of 40 m.
The beacon-to-glider performance, however, deteriorated due to missed contacts outside the duct, and was not described quantitatively.
In 2017, gliders were deployed in a region with no ice coverage \citep{Graupe2019}.
Ranges were linearly scaled by a statistical description of sound speed observations taken during the experiment, 1450 $\pm$ 6.5 m/s.
This resulted in an error of 550 m, which was reduced by a factor between 4 and 5, depending on the dive, using a post-processing acoustic arrival matching method.
Both cases exploit the lower duct for high fidelity communication at long ranges.
Unintuitively, the smaller scale nature of our deployment during ICEX20 is not a simplifying factor.
For source depths typical to vehicle operations, 30 to 200 m, the Beaufort Lens introduces a shadow zone that spans from 2 to 6 kilometers in range \citep{schmidt_acoustic_2016}.

Compared to previous small scale navigation efforts, the approach in this paper integrates real-time model-aided data processing to estimate a representative sound speed along a path from source to receiver, leveraging climatology, \textit{in situ} data, and fast acoustic modeling.
The paper is organized as follows.
Section \ref{sec:icex20} details the experimental approach and conditions during ICEX20.
Given that there is no GNSS ground truth for the vehicle position while underway, we first evaluate the real-time ranging performance of GPS-linked beacon-to-beacon communication events in section \ref{sec:realtime}.
Section \ref{sec:post} uses insights from field data to introduce a new ray filtering algorithm to improve range estimation.
Section \ref{sec:trilat} emulates the real-time processing pipeline to re-position beacon-to-beacon events and re-navigate AUV \emph{Macrura}.

@@

The results from this paper derive from data collected while deploying the AUV \emph{Macrura}, a custom Bluefin-21, during the Ice Exercise 2020 (ICEX20).
The experiment was conducted in the Beaufort Sea, from March 8th to 11th, at roughly 71.2\textdegree N.
The AUV deployment was supported by the Integrated Communication and Navigation Network (ICNN) \citep{schneider_self-adapting_2020,Randeni2020,randeni_high-resolution_2021} a specialized implementation of the LBL solution.
The ICNN was initially developed via numerous virtual experiments to ensure robust algorithms and interfaces between different hardware components. 
The simulation capabilities are largely physics-driven with a modular system of systems approach\textemdash an environmental simulator with sub-components for the ocean, including Arctic ice drift and ocean acoustic propagation; a vehicle simulator with sub-components for vehicle dynamics and navigation; a topside hardware simulator and acoustic communications simulator, both with a software-only configuration and a hardware-in-the-loop version \citep{schneider_netsim_2018}.
The virtual environment similarly emulates the interfaces between the real components to test the entire software pipeline.

@@

The ICNN is comprised of four ice buoys in a loose rectangle, roughly 2 km away from a central ice camp with a topside computer, as shown in Fig.~\ref{fig:icnnOverview}.
Each ice buoy is outfitted with a Garmin GPS 18x, with a pulse-per-second rising edge aligned to 1 microsecond and a spec sheet accuracy of 3 m, 95\% of the time.
They are also each equipped with a WHOI Micro-Modem \citep{Singh2006}, with a four-element receiver array, a single transmitter, and one-tenth of a millisecond resolution.
\llabel{1.6a} Acoustic messages were sent with a 10 kHz carrier frequency, 5 kHz bandwidth, and phase-shift keying (PSK) modulation on a time-division multiple access schedule with a thirty-second cycle, giving room for two-way communication throughout the mission volume.
Thus the ICNN is dependent on the successful decoding of acoustic transmissions.
The receive and transmit elements were split between shallow and deeper depths\textemdash 30 and 90 m\textemdash to provide better coverage across the shadow zone.
While each buoy only has one transmit depth, all buoys have both receive depths but the active receive layer is consistent across all buoys. \llabel{1.12}
The design of the ICNN enables a self-adapting network to transmit and receive at the optimal depth to maintain contact with the AUV \citep{schneider_self-adapting_2020}.
The buoys do not encompass the full horizontal range of the vehicle but are positioned to minimize overlap in trilateration for spherical positioning \citep{deffenbaugh1996posit}.

@@

In the face of GNSS-denied navigation, this approach was anecdotally successful, as shown in Fig. \ref{fig:vehicleRecovery}.
The AUV \emph{Macrura} was deployed through a hydrohole from an ice camp but recovered through an emergency hydrohole.
A random disk error stalled \emph{Macrura} underneath the ice but did not prevent it from transmitting its location.
Due to an incoming storm, a team drilled an exploratory hole at \emph{Macrura's} self-reported location, and were able to drill a separate hole 1 m away to tie the vehicle off to a physical marker on the ice.
Three days later, \emph{Macrura} was recovered\textemdash the ice camp had moved over 19 km.
The AUV's relative position to camp had also changed, from roughly 45\textdegree at 1000 m to 90\textdegree at 1100 m.
Drama aside, we view the emergency recovery as qualitative proof of the robustness of this navigation approach.
Nonetheless, this paper specifically addresses the third and fourth steps\textemdash the conversion of travel times into pseudoranges and its quantative effect on trilateration.
By focusing on pseudorange estimates between GPS-tracked beacons, and re-running the trilateration pipeline, the results are decoupled from all other mechanisms in the ICNN.

@@

An important component to our navigation solution is an accurate estimation of a representative SSP for the ocean volume.
Previous field experience, during the Ice Exercise 2016 (ICEX16), demonstrated the negative effects of the Beaufort Lens on tracking and communication \citep{schmidt_acoustic_2016}.
Fig. \ref{fig:sspExpectation} shows historical, modeled, and \textit{in situ} sound speed data for both ICEX16 and ICEX20.
These three input streams were selected to mirror the information available on a submarine (personal conversation with LT B. Howard and LT CDR D. Goodwin). \llabel{1.8}
In the field, the SSP information was shared with the vehicle via basis representation compression on a lightweight digital acoustic message \citep{bhatt_embedded_2022}. \llabel{2.7}
All modeled data comes from HYCOM \cite{chassignet_hycom_2007}, which does not seem to capture the forcing mechanisms that cause the Beaufort Lens.
For ICEX16, the data-driven profile was sourced from nearby Ice Tethered Profilers (ITP) after the field experiment \cite{Krishfield2008,toole_ice-tethered_2011} and exhibits a fairly deep lens; the historical profile is from the World Ocean Atlas.
For ICEX20, the chosen weights (data-driven) profile derives from initial CTD casts taken on site, showing an intense warm water intrusion; the baseline (historical) profile, showing moderate ducted conditions, comes from the average of March 2013 ITP data.
This month best matched sea ice and sound speed conditions at the beginning of ICEX20 \citep{bhatt_embedded_2022}.
It is important to note that all profiles that do show the Beaufort Lens do so with different local sound speed maxima at different depths, reflective of the wide range of lens properties observed for all ITP data in the region. \llabel{2.5}
The variability of the lens height and prominence is the main reason an updatable SSP was integrated into the ICNN solution.

@@

During ICEX20, the HYCOM profile was available but never deployed. \llabel{1.14}
For post-processing comparison, we introduce both the HYCOM profile and an isovelocity case, 1441.8 $\pm$ 3.7 m/s, as the mean and standard deviation of the observed sound speed profile over the first 200 m.
This is a contrived value taken in the style of \citet{Graupe2019} for the sake of comparison; the default value in the LAMSS simulator, which was not environmentally informed and used when no updates were available from the ICNN, was 1430 m/s.

@@

\subsection{Minimal bounce (MB) criteria}

\llabel{1.5} The fundamental challenge to implement GNSS-like navigation, especially in an acoustically complex propagation environment, is characterizing a single sound speed to compensate for the effects of ray refraction and reflection.
The use of the acoustic modem network for tracking relies on the accurate estimation of travel times between the submerged platform and LBL beacons, supported by clock synchronization and a pre-determined scheduling of acoustic events.
\llabel{2.9} For the Beaufort Lens in particular, the strong multipath effects make it virtually impossible to deterministically predict the modem's detected arrival time.
